---
title: "MATH 423 Final Project"
author: "Jeremy Kim - 260620525"
date: '2016-12-20'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
file<-"http://www.math.mcgill.ca/dstephens/Regression/Data/Project/nhanes-sub.csv"
data<-read.csv(file)


gender<-data$gender
age<-data$age
race3<-data$race3
educ<-data$educ
married<-data$married
hhinc<-data$hhinc
poverty<-data$poverty
homerooms<-data$homerooms
homeown<-data$homeown
weight<-data$weight
bmi<-data$bmi
pulse<-data$pulse
bpsys<-data$bpsys
bpdia<-data$bpdia
dirchol<-data$dirchol
totchol<-data$totchol
physact<-data$physact
alcday<-data$alcday
alcyear<-data$alcyear

n<-length(bpsys)

# model adequacy criteria
criteria.eval<-function(fit.obj,nv,bigsig.hat){
    cvec<-rep(0,5)
    SSRes<-sum(residuals(fit.obj)^2)
    p<-length(coef(fit.obj))
    cvec[1]<-summary(fit.obj)$r.squared
    cvec[2]<-summary(fit.obj)$adj.r.squared
    cvec[3]<-SSRes/bigsig.hat^2-n+2*p
    #AIC in R computes
    # n*log(sum(residuals(fit.obj)^2)/n)+2*(length(coef(fit.obj))+1)+n*log(2*pi)+n
    cvec[4]<-AIC(fit.obj)
    #BIC in R computes
    # n*log(sum(residuals(fit.obj)^2)/n)+log(n)*(length(coef(fit.obj))+1)+n*log(2*pi)+n
    cvec[5]<-BIC(fit.obj)
    return(cvec)
}
```

## bpdia

```{r bpdia}
bpdia.fit1<-lm(bpdia ~ age + alcday + alcyear + bmi + dirchol + educ + gender + hhinc + homeown + homerooms + married + physact + poverty+ pulse+ race3 + totchol + weight)
```
```{r bpdia_a, results="hide"}
# remove some effects
bpdia.fit2<-step(bpdia.fit1,test='F')
```
```{r bpdia_b}
anova(bpdia.fit1,bpdia.fit2,test='F')
```

Has high F value, thus predictors removed from fit 1 are not significant enough. Fit 2 is adequate.

Now add some second order interactions.
```{r bpdia2, results="hide"}
# add some second order interactions
bpdia.fit3<-step(bpdia.fit2,test='F', scope= list(upper=update(bpdia.fit2, ~.*.), lower= bpdia.fit2))
```
```{r bpdia2a}
anova(bpdia.fit2,bpdia.fit3, test='F')
# low p value fit 3 is adequate
```

Has low F value, thus interactions added are significant enough. Fit 3 is adequate.

## Model adequacy for bpdia

```{r bpdia_adeq, echo = FALSE}
bigs.hat<-summary(bpdia.fit3)$sigma
cvals<-matrix(0,nrow=3,ncol=5)
cvals[1,]<-criteria.eval(bpdia.fit1,n,bigs.hat)
cvals[2,]<-criteria.eval(bpdia.fit2,n,bigs.hat)
cvals[3,]<-criteria.eval(bpdia.fit3,n,bigs.hat)
Criteria<-data.frame(cvals)
names(Criteria)<-c('Rsq','Adj.Rsq','Cp','AIC','BIC')
rownames(Criteria)<-c('1', '2', '3')
round(Criteria,4)
```

Fit 3 has lowest AIC and highest \(R^2_{adj}\), thus it is most adequate.


## Residual plots for bpdia


```{r bpdia_plot, echo=FALSE}
par(mfrow=c(3,3),mar=c(4,2,1,2))
plot(age,residuals(bpdia.fit3),pch=19,cex=0.75, main="Age", xlab ="")
abline(h=0, lty=2)
plot(gender,residuals(bpdia.fit3),pch=19,cex=0.75, main = "Gender")
abline(h=0, lty=2)
plot(hhinc,residuals(bpdia.fit3),pch=19,cex=0.75,xaxt="n", main="Household Income", xlab ="")
axis(1, at=1:12, labels=levels(hhinc), las = 2, cex.axis = 0.8)
abline(h=0, lty=2)
plot(married,residuals(bpdia.fit3),pch=19,cex=0.75,xaxt="n" , main="Marital Status", xlab ="")
axis(1, at=1:6, labels=levels(married), las = 2, cex.axis = 0.8)
abline(h=0, lty=2)
plot(poverty,residuals(bpdia.fit3),pch=19,cex=0.75, main="Poverty")
abline(h=0, lty=2)
plot(pulse,residuals(bpdia.fit3),pch=19,cex=0.75, main="Pulse")
abline(h=0, lty=2)
plot(race3,residuals(bpdia.fit3),pch=19,cex=0.75,xaxt="n", main="Race", xlab = "")
axis(1, at=1:6, labels=levels(race3), las = 2, cex.axis = 0.8)
abline(h=0, lty=2)
plot(totchol,residuals(bpdia.fit3),pch=19,cex=0.75, main="Total HDL Cholesterol")
abline(h=0, lty=2)
plot(weight,residuals(bpdia.fit3),pch=19,cex=0.75, main="Weight")
abline(h=0, lty=2)


```
Residual plots seem reasonable with constant variance. Factor predictors have less consistent residuals but they seem to even out.

***
Diastolic blood pressure is higher for males (7.684), slightly lower for older people(-.4278), higher for less poor people (29.171, though offsetted by the interaction between poverty index and hhinc), lower for those with higher cholesterol (-7.197), highest for middle class (45k-75k hhinc) and lowest for 25k-45k hhinc, highest for other races and lowest for Blacks and Hispanics.  


## bpsys

```{r bpsys}
bpsys.fit1<-lm(bpsys ~ age + alcday + alcyear + bmi + dirchol + educ + gender + hhinc + homeown + homerooms + married + physact + poverty+ pulse+ race3 + totchol + weight)

```
```{r bpsys_a, results="hide"}
bpsys.fit2<-step(bpsys.fit1,test='F')

```
```{r bpsys_b}
anova(bpsys.fit1,bpsys.fit2,test='F')
```
Has high F value, thus predictors removed from fit 1 are not significant enough. Fit 2 is adequate.

Let's see if there are more predictors that can be dropped
```{r bpsys_c}

drop1(bpsys.fit2, test='F')
bpsys.fit3<-update(bpsys.fit2, ~. - dirchol - physact)
anova(bpsys.fit2,bpsys.fit3,test='F')
```
Has high F value, thus predictors removed from fit 2 are not significant enough. Fit 3 is adequate.
Now add some second order interactions.
```{r bpsys2}
# add some second order interactions
add1(bpsys.fit3,test='F', scope= update(bpsys.fit3, ~.*.))
bpsys.fit4<-update(bpsys.fit3, ~. +age:gender+age:poverty + age:race3 + age:totchol + alcday:bmi + alcday:educ + alcday:hhinc + alcday:poverty + alcday:pulse + alcday:race3 + educ:hhinc + educ:married + gender:hhinc + gender:married + hhinc:married + hhinc:race3 + homeown:married + married:race3 + married:totchol)
anova(bpsys.fit4,bpsys.fit3,test='F')
```
Has low F value, thus interactions added are significant enough. Fit 4 is adequate.

```{r bpsys2_a}
# drop some variable
drop1(bpsys.fit4, test='F')
bpsys.fit5<-update(bpsys.fit4, ~. - age:race3 - alcday:bmi - alcday:poverty - alcday:pulse - alcday:race3 )
anova(bpsys.fit5,bpsys.fit4,test='F')


```
Has high F value, thus predictors removed are not significant. Fit 5 is adequate.
```{r bpsys2_b}
# add some
add1(bpsys.fit5, test="F", scope = update(bpsys.fit3, ~.*.))
bpsys.fit6<- update(bpsys.fit5, ~. + alcyear:hhinc+alcyear:married +alcyear:totchol + educ:poverty + pulse:totchol)
anova(bpsys.fit5, bpsys.fit6, test='F')
```
Has low F value, thus predictors added are significant enough. Fit 6 is adequate.


## Model adequacy for bpsys

```{r bpsys_adeq, echo = FALSE}
bigs.hat<-summary(bpsys.fit6)$sigma
cvals<-matrix(0,nrow=6,ncol=5)
cvals[1,]<-criteria.eval(bpsys.fit1,n,bigs.hat)
cvals[2,]<-criteria.eval(bpsys.fit2,n,bigs.hat)
cvals[3,]<-criteria.eval(bpsys.fit3,n,bigs.hat)
cvals[4,]<-criteria.eval(bpsys.fit4,n,bigs.hat)
cvals[5,]<-criteria.eval(bpsys.fit5,n,bigs.hat)
cvals[6,]<-criteria.eval(bpsys.fit6,n,bigs.hat)
Criteria<-data.frame(cvals)
names(Criteria)<-c('Rsq','Adj.Rsq','Cp','AIC','BIC')
rownames(Criteria)<-c('1', '2', '3', '4', '5', '6')
round(Criteria,4)
```

Fit 6 has lowest AIC and highest \(R^2_{adj}\),, thus it is most adequate.

\newpage
## Residual plots for bpsys


```{r bpsys_plot, echo=FALSE}
par(mfrow=c(3,3),mar=c(4,2,1,2))
plot(age,residuals(bpsys.fit6),pch=19,cex=0.75, main="Age", xlab ="")
abline(h=0, lty=2)
plot(alcday,residuals(bpsys.fit6),pch=19,cex=0.75, main = "Drinks/Day")
abline(h=0, lty=2)
plot(alcyear,residuals(bpsys.fit6),pch=19,cex=0.75, main="Days drank/Year", xlab ="")
abline(h=0, lty=2)
plot(bmi,residuals(bpsys.fit6),pch=19,cex=0.75, main="BMI", xlab ="")

abline(h=0, lty=2)
plot(educ,residuals(bpsys.fit6),pch=19,cex=0.75, main="Education", xaxt="n")
axis(1, at=1:5, labels=levels(educ), las = 2, cex.axis = 0.8)
abline(h=0, lty=2)
plot(gender,residuals(bpsys.fit6),pch=19,cex=0.75, main="Gender")
abline(h=0, lty=2)
plot(hhinc,residuals(bpsys.fit6),pch=19,cex=0.75,xaxt="n", main="Household Income", xlab = "")
axis(1, at=1:12, labels=levels(hhinc), las = 2, cex.axis = 0.8)
abline(h=0, lty=2)
plot(homeown,residuals(bpsys.fit6),pch=19,cex=0.75, main="Home ownership")
abline(h=0, lty=2)
plot(married,residuals(bpsys.fit6),pch=19,cex=0.75,xaxt="n" , main="Marital Status", xlab ="")
axis(1, at=1:6, labels=levels(married), las = 2, cex.axis = 0.8)
abline(h=0, lty=2)
par(mfrow=c(2,2),mar=c(4,2,1,2))
plot(poverty,residuals(bpsys.fit6),pch=19,cex=0.75, main="Poverty")
abline(h=0, lty=2)
plot(pulse,residuals(bpsys.fit6),pch=19,cex=0.75, main="Pulse")
abline(h=0, lty=2)
plot(race3,residuals(bpsys.fit6),pch=19,cex=0.75,xaxt="n", main="Race", xlab = "")
axis(1, at=1:6, labels=levels(race3), las = 2, cex.axis = 0.8)
abline(h=0, lty=2)
plot(totchol,residuals(bpsys.fit6),pch=19,cex=0.75, main="Total HDL Cholesterol")
abline(h=0, lty=2)


```
Residual plots seem reasonable with constant variance. Factor predictors have less consistent residuals but they seem to even out. Drinks/day has higher variance for lower values but it seems to be caused by abundance of those who do not drink in a day.

***
Systolic blood pressure is slightly higher for older people, those who drink more, and those with higher BMI, highest for high school grads (59.051), highest for other races and lowest for Hispanics (incl. Mexicans) and whites, lower for less poor people (-1.3378). Lower and middle classes seems to have lowest bpsys, similar to bpdia. Those who live with their partner or are married have higher bpsys. Pulse has little effect on both blood pressures (-.273 and -.317 respectively). 